---
title: "DefeoES207_HW5"
author: "Shelby Defeo"
date: "5/2/2022"
output: html_document
---
1. Determine if a linear relation between drainage area and Q90 is present. Does a logarithmic transformation of the two variables improve the relation?
There is a very loosely linear relationship between the variables, but there is a strong grouping near the bottom and a scattered few points towards the right corner. The logarithmic transformation does appear to improve the relationship between the two variables. 
```{r}
#1. Determine if a linear relation between drainage area and Q90 is present. Does a logarithmic transformation of the two variables improve the relation?
##Load and organize data
load("q90_gagesinfo.RData")
print(q90_gagesinfo)
cfs <- q90_gagesinfo$Q90_cfs
area <- q90_gagesinfo$area_mi2

# Check linearity
plot(cfs,area)

#transform
logcfs <- log(cfs)
logarea <-log(area)
plot(logcfs, logarea)

```

2. Using the results from Question 1, create a regression equation that can be used to estimate Q90 based on drainage area. Is drainage area a significant predictor of the Q90 streamflow? Provide evidence. Does the sign of the coefficient on drainage area match your intuition? Explain.?
Drainage area is a significant predictor of stream flow because the p-value of 3.02e-16 is less than my chosen alpha of 0.05. The sign of the coefficient (positive) also matches my intuition because I had predicted that a larger drainage area would equate to larger stream flows.
General equation: y = B0 + B1X1 
Specific equation: y = -2.27734 + 1.17536 * area

Independent: Area
Dependent: Stream flow
```{r}
#2. Using the results from Question 1, create a regression equation that can be used to estimate Q90 based on drainage area. Is drainage area a significant predictor of the Q90 streamflow? Provide evidence. Does the sign of the coefficient on drainage area match your intuition? Explain.?
##Regression analysis
regstream1 <- lm(logcfs ~ logarea, q90_gagesinfo)
summary(regstream1)
par(mfrow=c(2,2))
        plot(regstream1)
```

3. Based on the plot of residuals from the regression analysis, does the assumption of homogeneous variance appear to be met? Why or why not? This will help us determine whether it is appropriate to use the regression equation to make predictions.
Homogeneity of variance does appear to be met because the points are clustered around the zero line, and there are approximately the same amount of points above and below.I will say that the data could be more tightly clustered and equally spread out, but this seems to fall within a "reasonable" deviation from perfect homogeneity of variance. Also looking at our data from the previous question, we see that the max and min are about equal distance from zero (0.88128 and -0.63567), and our median is near zero (-0.07835). 
```{r}
#3. Based on the plot of residuals from the regression analysis, does the assumption of homogeneous variance appear to be met? Why or why not? This will help us determine whether it is appropriate to use the regression equation to make predictions.
##Check homogenity of variance
par(mfrow=c(2,2))
        plot(regstream1)
```








4. Create a scatter plot of mortality rate versus level of nitric oxides. Does it appear that a linear model will fit these data well?
It appears that a linear model will not necessarily fit this data well. There is some pretty high and scattered variance in the graph.
```{r}
#4. Create a scatter plot of mortality rate versus level of nitric oxides. Does it appear that a linear model will fit these data well?
##Add in data
pollution <- read.csv("pollution.csv")
pollution

##Create a scatterplot
plot(pollution$mort, pollution$nox)
```

5. Conduct a regression analysis for mortality rate and level of nitric oxides. Interpret the slope coefficient for level of nitric oxides (including whether it is statistically significant) and the overall fit of the model. Include an evaluation of the residual plots from the regression as part of your analysis.
The slope coefficent for nitric oxides is -0.1039, indicating a negative relationship, and has a p-value of 0.557. This p-value is greater than the alpha I had set of 0.05 so it is not statistically significant. The residual standard error is high so the model is not a great fit for our data. The residuals plot from our regression shows data that is not clustered evenly or consistently around zero, and there is a lot of variation.
```{r}
#5. Conduct a regression analysis for mortality rate and level of nitric oxides. Interpret the slope coefficient for level of nitric oxides (including whether it is statistically significant) and the overall fit of the model. Include an evaluation of the residual plots from the regression as part of your analysis.
##Conduct linear regression
pol1 <- lm(pollution$mort ~ pollution$nox, pollution)
summary(pol1)

##Graph regression results
par(mfrow=c(2,2))
        plot(pol1)
```

6. Perform a logarithmic transformation on the data you analyzed in Part A. Conduct a regression analysis with the transformed data. Interpret the slope coefficient for nitric oxides. Include an evaluation of the residual plots from the regression as part of your analysis.
The slope coefficent is now positive for the logarithmic transformation of nitric oxides, indicating a positive relationship, and it is statistically significant with a p-value less than 0.05. The residual standard error is lower in this logarithmic transformation than the previous regression, meaning it is better fit to our dataset. The regression plot data is more clustered around zero and has less variation, but it still is not perfectly symmetrical and shows some variation.
```{r}
#6. Perform a logarithmic transformation on the data you analyzed in Part A. Conduct a regression analysis with the transformed data. Interpret the slope coefficient for nitric oxides. Include an evaluation of the residual plots from the regression as part of your analysis.
##Transforming data
logmort <- log(pollution$mort)

lognox <- log(pollution$nox)

pol2 <- lm(logmort ~ lognox, pollution)
summary(pol2)

##Graph regression results
par(mfrow=c(2,2))
        plot(pol2)
```

7. Now, let’s look at all pollutants simultaneously. Fit a model predicting mortality rate using levels of nitric oxides, sulfur dioxide, and hydrocarbons as inputs. Use logarithmic transformations if/when appropriate. Plot the fitted regression model and interpret the coefficients.
The model I ended up finding that seems to best fit the data invloved a logarithmic transformation for mortality, nitric oxides, and hydrocarbons, but not sulfur dioxide. The residual standard error of this model (pol6) is 0.05794 compared to the other versions I ran that were in the 50's. The coefficients of nitric oxide and sulfur dioxide are positive, indicating a positive relationship with mortality, while the hydrocarbon coefficient is negative, indicating a negative relationship with mortality. Both nitric oxides and hydrocarbons coefficients are statistically significant with an alpha of 0.05. In the residuals plot, the data are somewhat clustered around zero and have a similar plot to the last question. 
```{r}
#7. Now, let’s look at all pollutants simultaneously. Fit a model predicting mortality rate using levels of nitric oxides, sulfur dioxide, and hydrocarbons as inputs. Use logarithmic transformations if/when appropriate. Plot the fitted regression model and interpret the coefficients.

##Evaluate need for transformations
#sulfur dioxide
plot(pollution$so2, pollution$mort)
logs02 <- log(pollution$so2)
plot(logs02, logmort)
plot(logs02, pollution$mort)
plot(pollution$so2, logmort)

#hydrocarbons
plot(pollution$hc, pollution$mort)
loghc <- log(pollution$hc)
plot(loghc, logmort)
plot(loghc, pollution$mort)

##regression analysis comparisons
pol3 <- lm(pollution$mort ~ lognox + pollution$so2 + pollution$hc, pollution)
summary(pol3)

pol4 <- lm(pollution$mort ~ lognox + logs02 + pollution$hc, pollution)
summary(pol4)

pol5 <- lm(pollution$mort ~ lognox + pollution$so2 + loghc , pollution)
summary(pol5)

pol6 <- lm(logmort ~ lognox + pollution$so2 + loghc, pollution)
summary(pol6) #best model with lowest residual standard error

##Graph regression results
par(mfrow=c(2,2))
        plot(pol6)
```

BQ1. Investigate potential interaction effects among the three predictors. You can do so in a variety of ways (e.g., correlation tests, conditional plots, etc.). If you suspect interaction effects are important, refit the model with these interactions and interpret the fitted model coefficients.
Based on using the cor.test function, it does appear that there may be interaction between the variables. All of the correlation tests result in a p-value less than 0.05 so we can say that those variables are significantly correlated to each other. In the regression analysis, log nox has a p-value less than 0.05, indicating that there is correlation between log nox concentrations and mortality, and the coefficient there is positive. The interaction between log nox and log hc has a p-value less than 0.05 and a negative coefficient. This model is tell us that log nox is significantly correlated with mortality, and the relationship is positive, plus log nox and log hc have a significant interaction and the relationship is negative. Additionally, S02 pollution and loghc seem to have a positive interaction based on their intercept and low p-value. Loghc and s02 do not appear to be singificantly correlated with mortality based on their p-values being higher than my alpha of 0.05.
```{r}
#BQ1. Investigate potential interaction effects among the three predictors. You can do so in a variety of ways (e.g., correlation tests, conditional plots, etc.). If you suspect interaction effects are important, refit the model with these interactions and interpret the fitted model coefficients.
mort <- pollution$mort
s02 <-pollution$so2
hc <-pollution$hc
nox <- pollution$nox

##Investigate interactions
cor.test(lognox, s02, method=c("pearson"))
cor.test(lognox, loghc, method=c("pearson"))
cor.test(s02, loghc, method=c("pearson"))
cor.test(lognox, logmort, method=c("pearson"))
cor.test(loghc, logmort, method=c("pearson"))
cor.test(s02, logmort, method=c("pearson"))

##Regression
pol7 <- lm(logmort ~ lognox + pollution$so2 + loghc +lognox * pollution$so2 + lognox * loghc + loghc * pollution$so2, pollution)
summary(pol7) 
##Graph regression results
par(mfrow=c(2,2))
        plot(pol7)
        

```

BQ2. Of the different models you analyzed for this scenario, which seems to be most suitable for the data/analysis? Why?
Of the different models I analyzed for this scenario, this third and final one seems to be the most suitable. It had a low residual standard error of 0.04985, while the first logarithmic transformation model has a residual standard error of 0.06412 and the second model that included multiple variables had a residual standard error of 0.05794. This last model also had the highest adjusted R-squared of 0.4347, while the other models has R-squared of 0.06476 and 0.2363. The residuals on this last model were also pretty symmetrical around zero. 